---
title: "PS270 Final Project"
author: "Callie Fadness"
description: "My final project"
output: html_document
---

\newpage

## Introduction 

I am interested in exploring data related to... voting patterns around Wisconsin and median household income (Had to change after the broadband data was not working)

My research question is: How is median household income within a county in Wisconsin related to Republican presidential vote share in the state? My hypothesis is that Wisconsin counties with higher median household income will have lower Republican vote share. This expectation is supported by political communication research, which suggests that higher-income populations may have greater access to diverse news sources, broader social networks, and political discourse that can influence voting behavior, even in areas that might otherwise lean conservative. My independent variable will be the median household income in a Wisconsin county, measured in U.S. dollars using the most recent available estimates, which can vary widely between urban and rural counties. My dependent variable is the Republican presidential vote share in Wisconsin, measured as the percentage of the two-party vote received by the Republican candidate in each county for U.S. presidential elections. If my hypothesis is correct, I expect to observe a negative and statistically significant coefficient on median household income in a regression predicting Republican vote share. Evidence that would potentially disprove my hypothesis would include a statistically insignificant relationship or a positive relationship, indicating that higher income is not associated with lower Republican vote share at the county level in Wisconsin.

Loading Packages:
I am using data from Bea.gov (Bureau of Economic Analysis) and https://electionlab.mit.edu/data (MIT election data)
```{r}
library(tidyverse)
library(modelsummary)

ideology <- readr::read_csv("data/county_election.csv", show_col_types = FALSE)

wi_county_income <- read_csv("data/CAINC1_WI_1969_2023.csv", show_col_types = FALSE)

ideology_wi <- ideology |>
  filter(state_po == "WI")

ideology_wi |>
  head(10) |>
  knitr::kable(caption = "Preview of County Election Data (Wisconsin + Other States)")

wi_county_income |>
  head(10) |>
  knitr::kable(caption = "Preview of BEA County Income Data")
```

Data tidying 
```{r}
wi_income_clean <- wi_county_income |>
  select(GeoName, Description, `2023`) |>
  filter(Description == "Per capita personal income (dollars) 2/") |>
  mutate(
    income = as.numeric(`2023`),
    county_name = GeoName |>
      tolower() |>
      gsub(", wi", "", x = _) |>
      gsub(" county", "", x = _)
  ) |>
  select(county_name, income)

ideology_clean <- ideology_wi |>
  filter(state_po == "WI", party == "REPUBLICAN") |>
  mutate(county_name = tolower(county_name)) |>
  group_by(county_name) |>
  summarise(
    republican_share = sum(candidatevotes) / sum(totalvotes),
    .groups = "drop"
  )

wi_income_clean |>
  head(10) |>
  knitr::kable(
    caption = "WI County Income Data"
  )

ideology_clean |>
  head(10) |>
  knitr::kable(
    caption = "County Election Data with Republican Vote Share"
  )

```

## Data Section 

```{r}
income_visual <- ggplot(wi_income_clean, aes(x = income)) +
  geom_histogram(bins = 30, fill = "blue", color = "white") +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(30000, 100000)
  ) +
  labs(
    title = "Distribution of Median Household Income Across Wisconsin Counties",
    x = "Median Household Income (USD)",
    y = "Number of Counties"
  )
income_visual
```


##Results Section


*Project milestone 4: Used some LLM in this because joining data was difficult with my two data sources.*
This plot is my second visualization. I had to use LLM to use options(scipen = #) to get scientific notation away, along with using [1:n] because I needed to have a 1:1 ratio for each variable between data sets.My research question is: How is median household income within a county in Wisconsin related to Republican presidential vote share in the state? This visual on the graph helps see the relationship between both data sources (which was complex to code together into one). I can see that lower income around the 45k to 62k range tends to have a larger republican vote share within Wisconsin counties. 
```{r}
merged_data <- inner_join(wi_income_clean, ideology_clean, by = "county_name")

options(scipen = 999) 

ggplot(merged_data, aes(x = income, y = republican_share)) + 
  geom_point() + geom_smooth(method = "lm") + 
  labs( title = "Income vs Republican Vote Share (WI Counties)", 
        x = "Income", 
        y = "Republican Vote Share" )
```


**Regression** 
```{r}

merged_data <- merged_data |>
  mutate(income_10k = income / 10000)

model <- lm(republican_share ~ income_10k, data = merged_data)

modelsummary(model, stars = TRUE, fmt = 5)

broom::tidy(model) |>
  knitr::kable(
    digits = 3,
    caption = "Regression Results: WI Republican Vote Share on Median Household Income"
  )

```

  To view the relationship between median household income and Republican presidential vote share across Wisconsin counties, I created a scatter plot with a linear regression line. The plot shows a weak but slightly positive relationship between income and Republican vote share. Counties with higher median incomes tend to have somewhat higher Republican vote share on average, but there is still some observable variation across counties with similar income levels, particularly in the 45,000 to 65,000 income range. This suggests that income alone does not strongly predict voting patterns across Wisconsin.The regression results support this. The coefficient for income (measured in 10,000 dollar units) is positive, indicating that higher-income counties tend to have slightly higher Republican vote share. Specifically, a $10,000 increase in median household income is associated with an increase in Republican vote share equal to approximately 0.004 in the estimated model. However, this relationship is not statistically significant, suggesting we aren't able to confidently conclude that income is related to vote share in this data set.
  Overall, these results do not support my original hypothesis, which predicted that higher-income counties would have lower Republican vote share. Instead, the estimated relationship is weakly positive and statistically indistinguishable from zero. Even if interpreted substantively, the effect size is very small, meaning that differences in income correspond to only minor changes in predicted vote share. Finally, this analysis should not be interpreted causally. Income alone explains only a small portion of the variation in voting outcomes, suggesting that other factors—such as urban–rural differences, education levels, and Wisconsin’s status as a competitive “purple” state—are likely more important in shaping county-level political behavior.

## Conclusion  







