---
title: "PS270 Final Project"
author: "Callie Fadness"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

\newpage

## Project thoughts

I am interested in exploring data related to... voting patterns around Wisconsin and median household income (Had to change after the broadband data was not working)

My research question is: How is median household income within a county in Wisconsin related to Republican presidential vote share in the state? My hypothesis is that Wisconsin counties with higher median household income will have lower Republican vote share. This expectation is supported by political communication research, which suggests that higher-income populations may have greater access to diverse news sources, broader social networks, and political discourse that can influence voting behavior, even in areas that might otherwise lean conservative. My independent variable will be the median household income in a Wisconsin county, measured in U.S. dollars using the most recent available estimates, which can vary widely between urban and rural counties. My dependent variable is the Republican presidential vote share in Wisconsin, measured as the percentage of the two-party vote received by the Republican candidate in each county for U.S. presidential elections. If my hypothesis is correct, I expect to observe a negative and statistically significant coefficient on median household income in a regression predicting Republican vote share. Evidence that would potentially disprove my hypothesis would include a statistically insignificant relationship or a positive relationship, indicating that higher income is not associated with lower Republican vote share at the county level in Wisconsin.

Loading Packages:
I am using data from Bea.gov (Bureau of Economic Analysis) and https://electionlab.mit.edu/data (MIT election data)
```{r}
library(tidyverse)
library(dbplyr)
ideology <- read_csv("data/county_election.csv")
wi_county_income <- read_csv("data/CAINC1_WI_1969_2023.csv")
head(wi_county_income)
head(ideology)

```

Project milestone 3: data tidying 
```{r}
wi_income_clean <- wi_county_income |>
  select(GeoName, Description, `2023`) |>
  filter(Description == "Per capita personal income (dollars) 2/") |>
  filter(grepl(",", GeoName)) |>
mutate(
    income = as.numeric(`2023`),
    county_name = sub(" County,.*", "", GeoName)
  )

ideology_clean <- ideology |>
  select(state_po, county_name, candidate, candidatevotes, totalvotes)|>
  mutate(republican_share = candidatevotes / totalvotes)

wi_income_clean
ideology_clean

```

```{r}
income_visual <- ggplot(wi_income_clean, aes(x = income)) +
  geom_histogram(bins = 30, fill = "blue", color = "white") +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(30000, 100000)
  ) +
  labs(
    title = "Distribution of Median Household Income Across Wisconsin Counties",
    x = "Median Household Income (USD)",
    y = "Number of Counties"
  )
income_visual
```

*Project milestone 4: Used some LLM in this because joining data was difficult with my two data sources.*
This plot is my second visualization. I had to use LLM to use options(scipen = #) to get scientific notation away, along with using [1:n] because I needed to have a 1:1 ratio for each variable between data sets.My research question is: How is median household income within a county in Wisconsin related to Republican presidential vote share in the state? This visual on the graph helps see the relationship between both data sources (which was complex to code together into one). I can see that lower income around the 45k to 62k range tends to have a larger republican vote share within Wisconsin counties. 
```{r}

wi_income_clean <- wi_income_clean |>
  mutate(
    income = as.numeric(wi_income_clean$income),
    county_name = sub(" County,.*", "", GeoName)
  ) 
 
n <- min(length(wi_income_clean$income), length(ideology_clean$republican_share))

plot_data <- data.frame(
  income = wi_income_clean$income[1:n],
  rep_share = ideology_clean$republican_share[1:n]
)

options(scipen = 999)

ggplot(plot_data, aes(x = income, y = rep_share)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    title = "Income vs Republican Vote Share (WI Counties)",
    x = "Income",
    y = "Republican Vote Share"
  )
```






