---
title: "PS270 Final Project"
author: "Callie Fadness"
description: "My final project"
output: html_document
---

\newpage

## Exploring data related to voting patterns around Wisconsin and median household income 

**Hypothesis**
**How is median household income within a county in Wisconsin related to Republican presidential vote share in the state?** 

\newline

My hypothesis is that Wisconsin counties with higher median household income will have lower Republican vote share. This expectation is supported by primary elections that show specific areas and how they vote, suggesting that higher-income populations may have greater access to diverse news sources, educational facilities, and political discourse that can influence voting behavior, even in areas that might lean conservative. My independent variable will be the median household income in a Wisconsin county, measured in U.S. dollars using the most recent available estimates, which can vary between urban and rural counties. My dependent variable is the Republican presidential vote share in Wisconsin, measured as the percentage of the two-party vote received by the Republican candidate in each county for U.S. presidential elections. If my hypothesis is correct, I expect to observe a negative and statistically significant coefficient on median household income in a regression predicting Republican vote share. Evidence that would potentially disprove my hypothesis would include a statistically insignificant relationship or a positive relationship, indicating that higher income is not associated with lower Republican vote share at the county level in Wisconsin.

Loading Packages:
I am using data from Bea.gov (Bureau of Economic Analysis) and https://electionlab.mit.edu/data (MIT election data)
```{r, include=FALSE}

knitr::opts_chunk$set(message = FALSE,warning = FALSE)

library(tidyverse)
library(modelsummary)

```

```{r, echo=FALSE}

ideology <- readr::read_csv("data/county_election.csv", show_col_types = FALSE)

wi_county_income <- read_csv("data/CAINC1_WI_1969_2023.csv", show_col_types = FALSE)

ideology_wi <- ideology |>
  filter(state_po == "WI")

wi_income_clean <- wi_county_income |>
  select(GeoName, Description, `2023`) |>
  filter(Description == "Per capita personal income (dollars) 2/") |>
  mutate(
    income = as.numeric(`2023`),
    county_name = GeoName |>
      tolower() |>
      gsub(", wi", "", x = _) |>
      gsub(" county", "", x = _)
  ) |>
  select(county_name, income)

ideology_clean <- ideology_wi |>
  filter(state_po == "WI", party == "REPUBLICAN") |>
  mutate(county_name = tolower(county_name)) |>
  group_by(county_name) |>
  summarise(
    republican_share = sum(candidatevotes) / sum(totalvotes),
    .groups = "drop"
  )

wi_income_clean |>
  head(10) |>
  knitr::kable(
    col.names = c("WI County", "Income"),
    caption = "WI County Income Data"
  )

ideology_clean |>
  head(10) |>
  knitr::kable(
    col.names = c("WI County", "Republican Vote"),
    caption = "County Election Data with Republican Vote Share"
  )

```

## Data Section 

\newline

```{r, echo=FALSE}
income_visual <- ggplot(wi_income_clean, aes(x = income)) +
  geom_histogram(bins = 30, fill = "blue", color = "white") +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(30000, 100000)
  ) +
  labs(
    title = "Distribution of Median Household Income Across Wisconsin Counties",
    x = "Median Household Income (USD)",
    y = "Number of Counties"
  )
income_visual
```
\newline 

```{r, echo=FALSE}
ggplot(ideology_clean, aes(x = republican_share)) +
  geom_histogram(bins = 30, fill = "red", color = "white") +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(
    title = "Distribution of Republican Vote Share Across Wisconsin Counties",
    x = "WI Republican Vote Share",
    y = "WI Proportion of Counties"
  )
```
 

**Independent Variable: Distribution of Median Household Income Across WI Counties Analysis-**
\newline
The distribution of median household income across Wisconsin Counties is sourced from the Bureau of Economic Analysis as of 2023. This examines whether median household income in Wisconsin counties is related to Republican presidential vote share as of. The outcome variable is Republican vote share, measured as the proportion of total presidential votes in each county that went to the Republican candidate. The main predictor is median household income, measured in U.S. dollars using county-level income data. By combining these two data sets by county name, I can compare economic differences across counties with political voting patterns.
\newline

My hypothesis is that counties with higher median household incomes may have different levels of Republican vote share than lower income counties. To visualize and analyze this relationship, I use two visualizations and a linear regression model. If the slope coefficient for income is statistically significant (p-value less than 0.05), I would reject the null hypothesis of no relationship between income and Republican vote share. If the p-value is greater than 0.05, I fail to reject the null hypothesis, so income alone would not significantly predict Republican vote share.
\newline

The histogram above shows the distribution of median household income across Wisconsin counties. Most counties fall between roughly $45,000 and $65,000, with fewer counties at very low or very high income levels. This indicates that Wisconsin county incomes are mostly concentrated around the middle-income range, which is useful context before examining whether these differences are associated with voting behavior.

**Dependent Variable: Distributuion of Republican Vote Share Across WI Counties (2024) Analysis-**
The distribution of Republican vote share across Wisconsin is sourced from the MIT election data lab as of 2024. This histogram shows that there is a concentration of Republican votes in the 49-55% range. This shows that voters throughout counties in Wisconsin voted on average more Rpublican (in favor of Donald Trump). Overall, the distribution shows that Republican candidates, including Donald Trump in 2024, received relatively strong and consistent support across a majority of counties, with most falling near or slightly above the 50% threshold. This makes sense since Trump previously won Wisconsin in the 2024 general election. 

## Results Section 

```{r, echo=FALSE}
merged_data <- inner_join(wi_income_clean, ideology_clean, by = "county_name")

options(scipen = 999) 

ggplot(merged_data, aes(x = income, y = republican_share)) + 
  geom_point() + geom_smooth(method = "lm") + 
  labs( title = "Income vs Republican Vote Share (WI Counties)", 
        x = "Income", 
        y = "Republican Vote Share" )
```

**Regression** 
```{r, echo=FALSE}

merged_data <- merged_data |>
  mutate(income_10k = income / 10000)

model <- lm(republican_share ~ income_10k, data = merged_data)

modelsummary(model, stars = TRUE, fmt = 5)

broom::tidy(model) |>
  knitr::kable(
    digits = 3,
    caption = "Regression Results: WI Republican Vote Share on Median Household Income"
  )

```
**Scatter plot analysis:**
  To view the relationship between median household income and Republican presidential vote share across Wisconsin counties, I created a scatter plot with a linear regression line. The plot shows a weak but slightly positive relationship between income and Republican vote share. Counties with higher median incomes tend to have somewhat higher Republican vote share on average, but there is still some observable variation across counties with similar income levels, particularly in the 45,000 to 65,000 income range. This suggests that income alone does not strongly predict voting patterns across Wisconsin.The regression results support this. The coefficient for income (measured in 10,000 dollar units) is positive, indicating that higher-income counties tend to have slightly higher Republican vote share. Specifically, a $10,000 increase in median household income is associated with an increase in Republican vote share equal to approximately 0.004 in the estimated model. However, this relationship is not statistically significant, suggesting we aren't able to confidently conclude that income is related to vote share in this data set.
\newline

**Results and connection to hypothesis:**
  Overall, these results do not support my original hypothesis, which predicted that higher-income counties would have lower Republican vote share. Instead, the estimated relationship is weakly positive and statistically indistinguishable from zero. Even if interpreted substantively, the effect size is very small, meaning that differences in income correspond to only minor changes in predicted vote share. Finally, this analysis should not be interpreted causally. Income alone explains only a small portion of the variation in voting outcomes, suggesting that other factors—such as urban–rural differences, education levels, and Wisconsin’s status as a competitive “purple” state—are likely more important in shaping county-level political behavior.
\newline

**Standard error and p-values from this data:**
  This data regression shows under income_10k variable that there is a 0.01 standard error and a p-value of 0.671, meaning that this effect is not statistically significant, which means the observed relationship between both variables could be due to random chance instead of a true pattern in the data. As for the standard error, the estimated effect of income on Republican vote share is very small, but there is enough uncertainty that the real effect of this variable could have somewhat of an impact.
  Overall, this chart also shows that income alone does not appear to be a strong predictor of Republican vote share in Wisconsin counties.


## Conclusion  
In closing, I found little support for my hypothesis to prove an end result that median household income is strongly related to Republican presidential vote share across Wisconsin counties. Both the scatter plot and regression analysis showed only a very weak positive relationship. The main income coefficient (p-value) was not statistically significant. This suggests that differences in county income alone do not meaningfully explain variation in Republican voting patterns in Wisconsin. One major limitation of this analysis is that it only includes one predictor variable, which increases the risk of omitted confounding bias because other important factors—such as education levels, rural versus urban geography, racial demographics, or cultural influences—may better explain county voting behavior. In addition, this is observational county-level data, so the results from the election aren't correlated with the income that is reported, as we don't know who voted out of that sample. Data quality issues, missing values, or differences in how counties are measured across datasets could also have affected my findings, although I did my best to merge the sets to my best abilities. If I had more time and resources, I would improve the analysis by adding additional demographic and political variables, using a run through of multiple-variable regressions, and using a timeline of data across multiple election years to better understand how economic conditions interact with voting behavior over time.

**DISCLAIMER**

I have used LLM for this project in order to combine two very different and complex data sets for my project. I made sure to use it with integrity and ensure that I would not override my own work by using it. 

```{r print-code, ref.label=knitr::all_labels(), echo = TRUE, eval=FALSE}

```




